### 2025-09-18 paper/sec/introduction.tex
Progress on intro.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25}, on the other
hand, takes a more infrastructural approach to the transparency problem, linking visualisations to their data
sources directly. This allows visual outputs to support \emph{provenance queries}, user interactions (e.g.
mousing over a visual element) that reveal how visual features relate to data. The advantage of this approach
is that the interactions are provided automatically via trusted infrastructure, namely a query language or
general-purpose programming language which tracks dependencies as data flows through a computation. However,
these approaches are limited to outputs computed from data, such as visualisations. What is missing is a way
to extend these ``direct interrogation'' features to natural language itself, where the main claims of most
scholarly articles are actually made.

In this paper, we make the case that progress can be made on this problem by combining these two approaches.
We introduce the idea of \emph{transparent documents}, web-based scholarly articles where key fragments of
natural language are computed from data, like visualisations and other summaries, and that support interactive
provenance queries that reveal how key quantitative statements in the document relate to the data. We report
on a proof-of-concept LLM-based tool which supports a human-in-the-loop workflow for iteratively transforming
a preexisting opaque document into a transparent, data-driven counterpart. Our approach leverages the ability
of LLMs to understand technical language and synthesise code, and also the provenance-tracking runtime of an
open source programming language called Fluid, which we use as our target platform.

The following example, adapted from \cite{zhang18} via \cite{moosavi21}, shows how the language presented in
this paper can be used to achieve this \rpnote{Need to illustrate counterfactual/generalisability}:

\begin{figure}%[h]
    \centering
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-with-pointer.png}
    \vspace{1mm}
    \hrule
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-counterfactual-with-pointer.png}
    \caption{\emph{Transparent document} example, showing data-driven natural language (top
    vs.~bottom)}\label{fig:scigen-example-website}
\end{figure}

>>>>

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,bond25}, on the other hand,
takes a more infrastructural approach to the transparency problem, linking computed outputs to their data
sources directly. This allows visual outputs to support \emph{provenance queries}, user interactions (e.g.
mousing over a visual element) that reveal how visual features relate to data. The advantage of this approach
is that the interactions are provided automatically via trusted infrastructure, namely a query language or
general-purpose programming language which tracks dependencies as data flows through a computation. However,
these approaches are limited to outputs computed from data, such as visualisations. What is missing is a way
to extend these ``direct interrogation'' features to natural language itself, where the main claims of most
scholarly articles are actually made.

In this paper, we this gap by combining these two approaches, leveraging the ability of LLMs to understand
technical language and synthesise queries over data, and utilising the provenance-tracking infrastructure of
an open source programming language called Fluid (\url{https://f.luid.org/}) designed originally for
transparent visualisations~\citep{perera22,bond25}. We show how these two technologies can together support
the creation and deployment of \emph{transparent documents}, web-based scholarly articles where key
quantitative claims expressed in natural language --- such as the claim that one software system is faster
than another at solving a problem --- are computed from the relevant data, and that support interactive
provenance queries that reveal to a reader or reviewer the specific data elements that support the claim.

\figref{scigen-example-website}, adapted from \cite{zhang18}, shows how a transparent document appears to a
user.

\begin{figure}%[h]
    \centering
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-with-pointer.png}
    \vspace{1mm}
    \hrule
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-counterfactual-with-pointer.png}
    \caption{\emph{Transparent document} example, showing data-driven natural language (top
    vs.~bottom)}
    \label{fig:scigen-example-website}
\end{figure}

\rpnote{Need to illustrate counterfactual/generalisability}

\paragraph{Contributions}

Our specific contributions are:

\begin{itemize}
\item a proof-of-concept LLM-based tool for iteratively transforming a preexisting opaque document and
associated data set into a transparent, data-driven counterpart
\item an empirical evaluation of how well state-of-the-art models are able to solve the associated
interpretation and code synthesis problems
\end{itemize}

We leave implementing a full IDE authoring plugin (based on VSCode or Cursor, for example) for future work.

Our contribution is an agent-based tool which identifies text fragments which are candidates for being made
data-driven, generates the corresponding Fluid code, and then incrementally incorporates the code into a
data-driven version of the original document.
